Type-Directed, Whitespace-Delimited Parsing for Embedded DSLs

### 2013-06-14 globaldsl13/approach.tex
GlobalDSL13 final

\section{Approach}
\label{s:approach}

Figure~\ref{f:dsl-if} illustrates our approach to type-based extension for the example in Figure 1.  The class \verb|ComponentArch| defines a grammar extension to the language's expression grammar. This grammar must not conflict with the fixed base expression grammar, but because it is only used when an expression of the type \verb|ComponentArch| type is requested \emph{and} a whitespace-delimited block has been introduced, there is no need to guarantee that this grammar does not interfere with other type-associated grammars.

The declared grammar uses a standard BNF notation except for the declaration \verb|EXP of URL| on Line 7. This indicates that at this position in the grammar, a term of type \verb|URL| is needed. Thus, the domain-specific grammar (if any) associated with the \verb|URL| class may be used here. Because URLs are typically short, it may be undesirable to require a new whitespace-delimited block, so the grammar defines blocks delimited by brackets as equivalent to whitespace-delimited blocks as well, as can be seen being used in Figure 1.

\begin{figure}
  \centering\small
  \begin{lstlisting}[language=Python,morekeywords={grammar,where,ID,TYPE,EXP}]
class ComponentArch
  grammar ::= ((component|policy) \n)+ where 
    component ::= 
      ("external ")? "component " ID ":" TYPE
        [componentProperty*]?
    componentProperty ::= 
        "location "(EXP of URL)
      | "connects to" (ID ",")* ID
    ...
  \end{lstlisting}
  \caption{Type-Associated Grammar Definition}
  \label{f:dsl-if}
\end{figure}

%In this example, we first parse the if condition using a standard
%expression parser \texttt{StdExpParser}.  The expression parser is
%passed a sub-stream that contains only tokens preceding the end of
%the line.  After parsing the condition expression, we then accept an
%\texttt{INDENT} token, indicating that we are now expecting one or
%more indented lines.  This token is consumed by using a DSL for
%parsing, which is modeled after BNF syntax.  The definition of this
%parsing DSL is not shown, but it can be defined using the same approach.
%%described in this section.
%
%After that, we use a standard statement parser to parse the indented lines as
%a series of one or more statements and construct and return
%an \texttt{IfAST} object representing the if statement.  The
%\texttt{IfAST} class is an ordinary class that implements the
%\texttt{AST} interface, which requires the class to define its semantics via
%an interpreter or via translation to lower-level constructs.  The
%details of the AST interface are out of scope here.  Once the
%semantics of \keyw{if} are defined, we can use it as shown in
%the \texttt{abs} method, which computes the absolute value of its
%argument.
%
%
\paragraph{Parsing and Typechecking}
Wyvern source code is parsed using a standard whitespace-delimited lexer followed by an initial typechecking and parsing phase that occurs in tandem.  The lexer for Wyvern generates a stream of tokens structured hierarchically, where indentation (and the equivalent bracket notation) increases the depth of the stream. Indentation and de-indentation can thus be simply  understood as a special pair of tokens, \texttt{INDENT} and \texttt{DEDENT}, generated by the lexer and delimiting portions of the stream that will be handled in a special way during the next phase.

The standard Wyvern parser reads at a single level of the stream at a time. Initially, it parses the stream as a term of sort \verb|Exp|, which represents the core expressions and declaration forms of the language. This sort includes basic facilities for declaring and working with functions and variables, objects and perhaps other to-be-determined core data types. When an indented block is encountered, however, it is left in-place within the parsed syntax tree.

When the parsing phase at a particular level is complete, it is typechecked according to a largely conventional protocol. In Wyvern, the type of a function is determined by its declaration, and cannot be affected by its use (this constraint may be relaxed in some instances in future iterations of the language). Thus, when a function is being applied, the expected types of its arguments are known. If, at a particular location where a function argument was expected, an indented block is found, then it is parsed not according to the \verb|Exp| grammar, but according to the \emph{type-associated grammar} contained in that type's definition. Figure 4 shows an example of such a grammar. That grammar may contain productions of the form $\texttt{EXP of }\tau$ where $\tau$ is another type in scope. A delimited block must be present at this location as well, and it will be parsed by the type-associated grammar of $\tau$.

A particular function application can contain only one \emph{primary} indented block. If multiple arguments support DSL syntax, then all those other than the first in the argument ordering must be invoked by a partially-indented keyword. This can be seen in Figure \ref{f:dsl-multi}. Note that this partially-indented keyword ends the first delimited block and begins a new block for the keyword argument \verb|toServer| of the function \verb|db.sendQuery|. This can be seen as similar in form to keywords in functions in Smalltalk \cite{smalltalk}.

\paragraph{Keyword-Directed Invocation}
Keyword-directed invocation of a DSL is simply a special case of this approach. In particular, a keyword can be defined as a function with a single argument of a type specific to that keyword. The type contains the implementation of the domain-specific syntax associated with that keyword. If the construction of this type is controlled such that it can only be via the domain-specific syntax (and not any equivalent syntax in the core language), then it is precisely equivalent to keyword-directed invocation of a DSL.

As keyword-directed DSL invocation is highly explicit, we allow keywords to appear not just at the end of a line, but also within expressions.  In this case, the keyword determines how much of the expression (and potentially following indented lines) that it will parse.

@@

\begin{figure}
\centering
\begin{lstlisting}
db.sendQuery
    select x where x > 10
  toServer
    www.server.org
\end{lstlisting}
\caption{Function applications involving multiple white-space delimited arguments use partially-indented keywords.}
\label{f:dsl-multi}
\end{figure}
%\begin{figure}
%  \centering
%  \begin{lstlisting}
%class Database class implements Parser
%	class meth parse(db : AST,
%	                 stream : TokStream):AST
%		if (stream ::= "where")
%			// parse where clause
%		else ... // parse other constructs
%	... // other definitions for Database
%
%// equivalent code for if using types
%class IfKeyType class implements Parser
%	class meth parse(ifKeyword : AST,
%	                 stream:ParseStream):AST
%		val cond = ... // same code as previous Figure
%	class prop instance = new IfKeyType // singleton
%val (@if@) = IfKeyType.instance
%  \end{lstlisting}
%  \caption{Type-based DSL definition}
%  \label{f:dsl-define}
%\end{figure}
%
\paragraph{Procedural Parsing}
The grammar definition in Figure 4 is declarative and relies on a simple parser generation framework included in Wyvern. It may be desirable for more sophisticated parsing algorithms to be encoded for particular domain-specific languages. For example, an interoperability layer with another full-scale programming language may wish to support parsing features not expressible in Wyvern's declarative grammar language. It can be observed, however, that a declarative grammar inside a class definition can be thought of as inducing a class method (a concept also borrowed from Smalltalk \cite{smalltalk}) called \verb|parse|, that transforms a stream of tokens to an AST representation. This interface can be exposed directly to programmers who wish to extend the parsing capabilities of the language in more sophisticated ways. We leave the details of the AST representation and its processing during later stages of compilation as future work.

>>>>

% !TEX root = globaldsl13.tex
\section{Approach}
\label{s:approach}
The examples above demonstrate the core mechanism used in Wyvern: each expression or declaration can contain at most one tilde (\verb|~|). The line following the term containing the tilde must begin an indented block. This block will be parsed according to a grammar associated with the expected type of the expression where the tilde occurred. In Figure \ref{f:dsl-arch}, this type was determined by an explicit type annotation. In Figures \ref{f:dsl-query} and \ref{f:dsl-presentation}, it was determined implicitly by the argument types of the function being called. Although a single tilde per expression may initially seem limiting, we can see in Figure 3 that the use of a \keyw{where} clause allows for the use of multiple DSLs within a single expression.

Figure~\ref{f:dsl-if} shows how users equip types with domain-specific grammars, here for the Architecture example in Figure \ref{f:dsl-arch}. The grammar associated with \lstinline{Architecture} is defined at the top level by a specially-named production, \keyw{grammar}, and it also defines two sub-productions for component and policy specifications. Each component specification includes a name, a type and an optional list of attributes. The name is specified using the \keyw{ID} production, which is a globally-available production that matches valid Wyvern (i.e. the GPL's) identifiers. Similarly, the \keyw{TYPE} production matches Wyvern types. These productions cannot be extended directly, but can be used within DSLs as needed and thus conflicts are detected when the library is compiled, rather than deferred to link-time.

One component attribute in the Architecture DSL is \lstinline{location}. On line \ref{f:dsl-if}.6, we see the use of a production defined in another type, \verb|URL|. This means that only URL literals are valid at that position, but \emph{not} any Wyvern expression of type \verb|URL|. To instead ask for any Wyvern expression of a particular type, we use the notation used in the \lstinline{policy} production on line \ref{f:dsl-if}.8. Here, a named policy is defined as an identifier followed by an equals sign and a Wyvern expression of type \lstinline{Policy}. This is denoted by the form \keyw{EXP}\verb| : T|, where \verb|T| is a type in scope of the grammar definition. Unlike in the URL example above, if there is a DSL associated with this type, it can only be used within a whitespace-delimited block introduced by a tilde. This key distinction can be seen in Figure \ref{f:dsl-arch}. Again, it can be modularly verified that this grammar does not contain conflicts, assuming that the core Wyvern grammar does not change. We expect it to become stable relatively early in its development, with most new features introduced via the embedded DSL mechanism.

\begin{figure}
  \centering\small
  \begin{lstlisting}[language=Python,morekeywords={type,grammar,where,ID,TYPE,EXP}]
type Architecture
  grammar ::= (component|policy)+ 
  component ::= "external"? "component" 
                     ID ":" TYPE 
                    ((componentAttr)*)?
  componentAttr ::= "location " URL.grammar
                       | "connects to" (ID ",")* ID
  policy ::= "policy" ID "=" (EXP : Policy)
  \end{lstlisting}
  \caption{Type-Associated Grammar}
  \label{f:dsl-if}
\end{figure}

%In this example, we first parse the if condition using a standard
%expression parser \texttt{StdExpParser}.  The expression parser is
%passed a sub-stream that contains only tokens preceding the end of
%the line.  After parsing the condition expression, we then accept an
%\texttt{INDENT} token, indicating that we are now expecting one or
%more indented lines.  This token is consumed by using a DSL for
%parsing, which is modeled after BNF syntax.  The definition of this
%parsing DSL is not shown, but it can be defined using the same approach.
%%described in this section.
%
%After that, we use a standard statement parser to parse the indented lines as
%a series of one or more statements and construct and return
%an \texttt{IfAST} object representing the if statement.  The
%\texttt{IfAST} class is an ordinary class that implements the
%\texttt{AST} interface, which requires the class to define its semantics via
%an interpreter or via translation to lower-level constructs.  The
%details of the AST interface are out of scope here.  Once the
%semantics of \keyw{if} are defined, we can use it as shown in
%the \texttt{abs} method, which computes the absolute value of its
%argument.
%
%
\paragraph{Parsing and Typechecking}
Wyvern source code is parsed in two phases. The first phase uses a standard declarative whitespace-delimited parser (e.g. \cite{rule2013principled}) where all whitespace-delimited blocks are left as unparsed ``DSL literals''. The second phase occurs in tandem with typechecking. When a tilde is encountered, the compiler determines the expected type where the tilde appeared (based on function signatures, or explicit type annotations) and parses the subsequent whitespace-delimited block according to the grammar associated with that type. The baseline indentation is stripped from this block, so it appears to the parser as if the DSL begins on the leftmost column of the block. Any Wyvern expressions that occur internally to a DSL (such as the policy specification in Figure \ref{f:dsl-arch} or curly-brace-delimited expressions in Figure \ref{f:dsl-presentation}) are also parsed and typechecked at this time. After parsing a DSL block, it must be verified and translated to Wyvern code. The mechanism for doing this is similar, at a high level, to that for defining the grammar itself, but the details of these subsequent steps are beyond the scope of this paper.

\paragraph{Procedural Parsing}
The grammar definition in Figure 4 is declarative and relies on a parser generator included as part of Wyvern. It may be desirable in some cases, such as when a grammar cannot easily be expressed within our declarative framework, or when an existing parser can be called into, to allow a parsing algorithms to be encoded directly. An important use case is for interoperability layers between Wyvern and other full-scale programming languages, particularly ones for which parsing is known to be difficult (e.g. C, Haskell, Python).

It can be observed that a declarative grammar inside a class definition can be seen as inducing a class method (that is, an operation defined on the class itself, which can be invoked by the compiler or run-time system -- a concept borrowed from Smalltalk \cite{smalltalk}) called \lstinline|parse|, that transforms a string to some AST representation. This lower-level interface is exposed directly to programmers who wish to specify parsing in a procedural manner.

%using a standard whitespace-delimited lexer followed by an initial typechecking and parsing phase that occurs in tandem.  The lexer for Wyvern generates a stream of tokens structured hierarchically, where indentation (and the equivalent bracket notation) increases the depth of the stream. Indentation and de-indentation can thus be simply  understood as a special pair of tokens, \texttt{INDENT} and \texttt{DEDENT}, generated by the lexer and delimiting portions of the stream that will be handled in a special way during the next phase.
%
%The standard Wyvern parser reads at a single level of the stream at a time. Initially, it parses the stream as a term of sort \verb|Exp|, which represents the core expressions and declaration forms of the language. This sort includes basic facilities for declaring and working with functions and variables, objects and perhaps other to-be-determined core data types. When an indented block is encountered, however, it is left in-place within the parsed syntax tree.

%When the parsing phase at a particular level is complete, it is typechecked according to a largely conventional protocol. 
%In Wyvern, the type of a function is determined by its declaration, and cannot be affected by its use (this constraint may be relaxed in some instances in future iterations of the language). Thus, when a function is being applied, the expected types of its arguments are known. If, at a particular location where a function argument was expected, an indented block is found, then it is parsed not according to the \verb|Exp| grammar, but according to the \emph{type-associated grammar} contained in that type's definition. Figure 4 shows an example of such a grammar. That grammar may contain productions of the form $\texttt{EXP of }\tau$ where $\tau$ is another type in scope. A delimited block must be present at this location as well, and it will be parsed by the type-associated grammar of $\tau$.
%
%A particular function application can contain only one \emph{primary} indented block. If multiple arguments support DSL syntax, then all those other than the first in the argument ordering must be invoked by a partially-indented keyword. This can be seen in Figure \ref{f:dsl-multi}. Note that this partially-indented keyword ends the first delimited block and begins a new block for the keyword argument \verb|toServer| of the function \verb|db.sendQuery|. This can be seen as similar in form to keywords in functions in Smalltalk \cite{smalltalk}.

@@

%\begin{figure}
%\centering
%\begin{lstlisting}
%db.sendQuery
%    select x where x > 10
%  toServer
%    www.server.org
%\end{lstlisting}
%\caption{Function applications involving multiple white-space delimited arguments use partially-indented keywords.}
%\label{f:dsl-multi}
%\end{figure}
%\begin{figure}
%  \centering
%  \begin{lstlisting}
%class Database class implements Parser
%	class meth parse(db : AST,
%	                 stream : TokStream):AST
%		if (stream ::= "where")
%			// parse where clause
%		else ... // parse other constructs
%	... // other definitions for Database
%
%// equivalent code for if using types
%class IfKeyType class implements Parser
%	class meth parse(ifKeyword : AST,
%	                 stream:ParseStream):AST
%		val cond = ... // same code as previous Figure
%	class prop instance = new IfKeyType // singleton
%val (@if@) = IfKeyType.instance
%  \end{lstlisting}
%  \caption{Type-based DSL definition}
%  \label{f:dsl-define}
%\end{figure}
%

### 2013-06-14 globaldsl13/conclusion.tex
GlobalDSL13 final

\section{Conclusion} % and Future Work}
\label{s:conclusion}

In this paper, we described how extensible parsing in Wyvern makes for
a solid platform to support whitespace-delimited, type-directed embedded DSLs. In the
future, we aim to implement a wide variety of DSLs in Wyvern tweaking
our approach and implementation thereof to provide a comprehensive example of
supporting multiple interacting DSLs in a safe and easy-to-use manner.

% \todo{tie features to goals}

@@

>>>>

% !TEX root = globaldsl13.tex
%\section{Conclusion} % and Future Work}
%\label{s:conclusion}
%
%In this paper, we described how extensible parsing in Wyvern makes for
%a solid platform to support whitespace-delimited, type-directed embedded DSLs. In the
%future, we aim to implement a wide variety of DSLs in Wyvern tweaking
%our approach and implementation thereof to provide a comprehensive example of
%supporting multiple interacting DSLs in a safe and easy-to-use manner.
%
% \todo{tie features to goals}

@@

\section*{Acknowledgements}
We thank the anonymous reviewers for helpful comments, and acknowledge the support of the Department of Defense and the Air Force Research Laboratory. CO is supported by the NSF Graduate Research Fellowship.

### 2013-06-14 globaldsl13/globaldsl13.tex
GlobalDSL13 final

\documentclass[english]{sigplanconf}
\usepackage{palatino}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{listings}
\usepackage{url}
\usepackage{babel}
\usepackage{graphicx}
%\usepackage{microtype}

@@

\lstset{tabsize=3, basicstyle=\ttfamily\small, commentstyle=\itshape\rmfamily, numbers=left, numberstyle=\tiny, language=java,moredelim=[il][\sffamily]{?},mathescape=true,showspaces=false,showstringspaces=false,columns=fullflexible,xleftmargin=15pt,escapeinside={(@}{@)}, morekeywords=[1]{application,policy,external,component,connects,to,meth,let,in,val,where,return,group,by,within,connect,with,attr,html,head,title,style,body,div,keyword,location,schema,must,salt,select,text}}
\lstloadlanguages{Java,VBScript,XML,HTML,Python}

@@

@@

\title{Whitespace-Delimited, Type-Directed Parsing for Embedded DSLs}

\authorinfo{Cyrus Omar, Benjamin Chung$^{1}$, Darya Kurilova, Alex Potanin$^{2}$,
  and Jonathan Aldrich}
{Carnegie Mellon University}
{\{comar, darya, aldrich\}@cs.cmu.edu, bwchung@andrew.cmu.edu$^{1}$, and alex@ecs.vuw.ac.nz$^{2}$}
% Maybe just keep my cs account active after I leave for a few years?
% Otherwise: alex@ecs.vuw.ac.nz is my email...
\maketitle

@@

Domain-specific languages enhance ease of use, expressiveness and
verifiability, 
but defining and using different 
DSLs within a single application remains difficult.  
We present an approach to embedded DSLs where:
1) a whitespace-delimited lexer is shared across DSLs, and 
2) the parsing and type checking phases occur in tandem so that types can define domain-specific parsing strategies governing whitespace-delimited blocks.
We argue that this approach occupies a sweet spot with both 
high expressiveness and ease of use. We outline a design
and ongoing implementation of this strategy
and examine the relationship between the type-directed and keyword-directed strategies for parsing of embedded DSLs.

\end{abstract}

\begin{sloppypar}

@@

\end{sloppypar}

\bibliographystyle{plain}
\bibliography{biblio}
\end{document}

>>>>

%\documentclass[english]{sigplanconf}
\documentclass[english]{sig-alternate}
\usepackage{times}
%\usepackage[T1]{fontenc}
%\usepackage[latin9]{inputenc}
\usepackage{listings}
\usepackage{url}
\usepackage{babel}
\usepackage{graphicx}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\renewcommand{\ttdefault}{pcr}
%\usepackage{microtype}

@@

\lstset{tabsize=2, basicstyle=\ttfamily\small, commentstyle=\itshape\rmfamily, numbers=left, numberstyle=\color{gray}\ttfamily\tiny, language=java,moredelim=[il][\sffamily]{?},mathescape=true,showspaces=false,showstringspaces=false,columns=fullflexible,xleftmargin=15pt,escapeinside={(@}{@)}, morekeywords=[1]{application,policy,external,component,connects,to,meth,let,in,val,where,return,group,by,within,connect,with,attr,html,head,title,style,body,div,keyword,location,schema,must,salt,select,text,type,then}}
\lstloadlanguages{Java,VBScript,XML,HTML,Python}

@@

   \conferenceinfo{GlobalDSL}{'13, Montpellier, France}
   \CopyrightYear{2013}
   \crdata{978-1-4503-2043-6}

@@

\title{Type-Directed, Whitespace-Delimited Parsing for Embedded DSLs}
\numberofauthors{1} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
% 1st. author
\alignauthor
Cyrus Omar, Benjamin Chung, Darya Kurilova, Alex Potanin$^{1}$ and Jonathan Aldrich\\
       \affaddr{Carnegie Mellon University}\\
%       \affaddr{1932 Wallamaloo Lane}\\
%       \affaddr{Wallamaloo, New Zealand}\\
       \email{\{comar, bwchung, darya, aldrich\}@cs.cmu.edu, and alex@ecs.vuw.ac.nz$^{1}$}
% 2nd. author
%\alignauthor
%G.K.M. Tobin\\
%       \affaddr{Institute for Clarity in Documentation}\\
%       \affaddr{P.O. Box 1212}\\
%       \affaddr{Dublin, Ohio 43017-6221}\\
%       \email{webmaster@marysville-ohio.com}
% 3rd. author
%\alignauthor Lars Th{\o}rv{\"a}ld\\
%       \affaddr{The Th{\o}rv{\"a}ld Group}\\
%       \affaddr{1 Th{\o}rv{\"a}ld Circle}\\
%       \affaddr{Hekla, Iceland}\\
%       \email{larst@affiliation.org}
}

%\authorinfo{Cyrus Omar, Benjamin Chung, Darya Kurilova, Alex Potanin$^{1}$ 
%  and Jonathan Aldrich}
%{Carnegie Mellon University}
%{\{comar, bwchung, darya, aldrich\}@cs.cmu.edu, and alex@ecs.vuw.ac.nz$^{1}$}
% Maybe just keep my cs account active after I leave for a few years?
% Otherwise: alex@ecs.vuw.ac.nz is my email...
\maketitle

@@

Domain-specific languages improve ease-of-use, expressiveness and
verifiability, 
but defining and using different 
DSLs within a single application remains difficult.  
We introduce an approach for embedded DSLs where 1) whitespace delimits DSL-governed blocks, and 2) the parsing and type checking phases occur in tandem so that the expected type of the block determines which domain-specific parser governs that block.
We argue that this approach occupies a sweet spot, providing   
high expressiveness and ease-of-use while maintaining safe composability. We introduce the design, provide examples and describe an ongoing implementation of this strategy in the Wyvern programming language. We also discuss how a more conventional keyword-directed strategy for parsing of DSLs can arise as a special case of this type-directed strategy.
\end{abstract}
\begin{sloppypar}

@@

\bibliographystyle{abbrv}
\bibliography{biblio}
\end{sloppypar}

\end{document}

### 2013-06-14 globaldsl13/intro.tex
GlobalDSL13 final

\section{Introduction}
\label{s:intro}

Domain-specific languages (DSLs) \cite{shellchapter, coldfusion,
  sinatra}
% have been widely-studied because they 
enable the expression of  
domain-specific ideas naturally and permit domain-specific
verification and compilation strategies.  For DSLs to reach their full
potential, however, it must be possible to easily use multiple
DSLs at once within a host general-purpose language (GPL), such that DSL code
and GPL glue code can inter-operate to form a complete application. More specifically, we propose the following essential design criteria:

@@

\item \emph {Composability}: It should be possible to use multiple DSLs and a GPL
%, in addition to a GPL,
within a single program unit.  %Here 
Within the text file-based paradigm familiar to most developers, this 
means including multiple DSLs within a single file.
 Moreover, it should be possible to embed code written in one DSL
  within another DSL without them requiring knowledge of each other or interfering with one another. So, more specifically, DSLs should be \emph{safely composable}.

\item \emph{Interoperability}: DSLs should be able to \emph{interoperate}. A necessary condition for interoperability is that it should be
  possible to pass and operate on values 
%such as functions or data structures
  defined in foreign DSLs. Additional requirements, such as the ability to do so naturally (rather than requiring a large amount of ``glue code'') and safely may also be relevant in many settings. 
%Moreover,
  Minimally, however, if one DSL defines a function or data structure satisfying an interface specified in
  a common interface description language (such as the type system of
  a host GPL), code written in another DSL should be able to use those
  values according to that interface, without requiring that the client DSL
  have knowledge of the details of the provider DSL or \emph{vice versa}.

\end{itemize}

@@

In addition to composability and interoperability, we believe that to be most useful, a system supporting DSLs should have the following characteristics:

\item \emph{Flexibility}: Have flexible enough DSL syntax to support a wide variety of text-based notations;

\item \emph{Identifiability}: Make it easy for programmers to identify which code is written in which DSL;

\item \emph{Consistency}: Encourage DSLs written in similar styles in order to enhance readability and learnability of each DSL;

\item \emph{Simplicity}: Keep the cost of defining a DSL low.

%\item Share conventions between DSLs and a host language, making each DSL easier for programmers to learn, helping programmers to identify which code is written in which DSL, and avoiding unintended conflicts between DSLs. \todo{Avoid conflicts (visual and real), enhance learnability}

%\item Reuse low-level mechanisms and design decisions from a host language, thereby reducing the cost of defining DSLs.  \todo{Easy to define DSLs}

\end{itemize}

We are developing a comprehensive approach that has the potential to meet these design requirements well, based on defining DSLs within a language specifically designed for extensibility from within.
The approach inherently supports multiple DSLs in the same file, and
supports interoperability because the DSLs are defined by translation
into a common internal language.  One key aspect of our proposed solution, and one of the focal points of this paper, is
that we restrict the syntax of DSLs in a few key ways that greatly simplify
the mechanism without significantly decreasing expressiveness:
\begin{itemize}

\item The host language and its DSLs share a tokenization and lexing 
  strategy, standardizing conventions for identifiers, operators,
  constants, and comments.  This avoids the cost of defining lexing
  within each DSL, avoids many kinds of low-level clashes between
  languages, and makes the composed language more readable. Note that this does not limit the ability of a DSL to define new keywords and other constructs.

\item Constructs in the host language and its DSLs are
  \emph{whitespace-delimited} at the top level, according to a particular strategy that we will describe.  Various forms of parentheses 
  can also serve as delimiters.  Thus, indentation levels or
  parenthesized expressions clearly delimit blocks that are governed by a particular 
  language.  This makes the boundaries of each DSL clear to
  the programmer and the compiler, enhancing usability and guaranteeing that 
  subtle conflicts cannot arise.

\end{itemize}

Within this syntactic framework, we then propose a novel syntax extension
mechanism in which the \emph{expected type of an expression}, rather than an explicit keyword, 
%, rather than a keyword, 
signals the switch to a new DSL that parses the tokenized and lexed code within a subsequent delimited block according to a type-associated grammar. The  more common keyword-directed strategy arises as a special case of this type-directed strategy. 
We will begin with some motivating examples (\S\ref{s:motivation}), describe our approach (\S\ref{s:approach}) and implementation (\S\ref{s:implementation}), and conclude with related work (\S\ref{s:related}).

>>>>

% !TEX root = globaldsl13.tex
\section{Introduction}
\label{s:intro}

Domain-specific languages (DSLs) \cite{fowler2010domain} % have been widely-studied because they 
allow developers to work with   
specialized abstractions in a natural manner, and allow for specialized 
verification and compilation strategies that can improve verifiability and performance. However, for DSLs to reach their full
potential, it must be simple to define a new DSL, invoke it when needed, and to use multiple
DSLs within a host general-purpose language (GPL), such that pieces of DSL code
 can interoperate to form a complete application. These intuitions are captured by the following core design criteria that govern our work:

@@

\item \emph {Composability}: It should be possible to use multiple DSLs and a GPL
%, in addition to a GPL,
within a single program unit.  %Here 
Within the file-based paradigm used by most contemporary languages, this 
means including multiple DSLs within a single file.
 Moreover, it should be possible to embed code written in one DSL
  within another DSL when appropriate, without requiring them to have specific knowledge of each other. This should be possible without interference between DSLs used in any combination: DSLs should be \emph{safely composable}.

\item \emph{Interoperability}: It should be
  possible to pass around and operate on values 
%such as functions or data structures
  that were defined in foreign DSLs in a reasonably natural manner (that is, without requiring large amounts of ``glue code''). Additional requirements, such as the ability to do so with the safety guarantees provided in the foreign DSL, may also be relevant in many settings.

%Moreover,
  %Minimally, DSLs should be able to define a function or data structure that satisfies an interface specified in
  %a common interface description language (such as the type system of
  %a host GPL), code written in another DSL should be able to use those
  %values according to that interface, without requiring that the client DSL
  %have knowledge of the details of the provider DSL or \emph{vice versa}.
\end{itemize}

@@

In addition to these fundamental criteria, we believe that to be most useful, a system supporting DSLs should satisfy the following related design criteria:

\begin{itemize}
\item \emph{Flexibility}: Support a variety of notations and new language mechanisms, with minimal bias;
\item \emph{Modularity}: Support defining DSLs as combinations of reusable components distributed directly within  libraries;
\item \emph{Identifiability}: Make it easy for programmers to identify which code is written in which DSL and what it means;
%\item \emph{Consistency}: Encourage DSLs written in similar styles whenever possible in order to enhance readability and learnability of each DSL;
\item \emph{Simplicity}: Keep the complexity and cost of both defining and invoking a DSL as low as possible.
%\item Share conventions between DSLs and a host language, making each DSL easier for programmers to learn, helping programmers to identify which code is written in which DSL, and avoiding unintended conflicts between DSLs. \todo{Avoid conflicts (visual and real), enhance learnability}
%\item Reuse low-level mechanisms and design decisions from a host language, thereby reducing the cost of defining DSLs.  \todo{Easy to define DSLs}
\end{itemize}

We are developing a comprehensive language design, \emph{Wyvern}, that we believe can satisfy these design criteria well, and that specifically considers language-internal extensibility from the start. In Wyvern, DSL developers define the run-time semantics of DSL constructs via translation
into a common host language, as in many other DSL frameworks. The novelty of the proposed extensibility mechanism lies in the ways in which we delimit and determine the \emph{scope} of DSL code:
\begin{itemize}
%\item The host language and its DSLs share a tokenization and lexing 
%  strategy, standardizing conventions for identifiers, operators,
%  constants, and comments.  This avoids the cost of defining lexing
%  within each DSL, avoids many kinds of low-level clashes between
%  languages, and makes the composed language more readable. Note that this does not limit the ability of a DSL to define new keywords and other constructs.
% 
\item Wyvern is a \emph{whitespace-delimited} language. Source code that is governed by a DSL, rather than the GPL, occurs in whitespace-delimited blocks and must be indented further than the GPL line introducing it. A decrease in indentation relative to the baseline of the DSL block signals its end. This scheme delimits the scope of each DSL in a clear manner, both to 
  the programmer and the top-level parser, supporting the principle of identifiability.  It also allows Wyvern to avoid restrictions on a DSL's use of delimiters internally. Because the GPL grammar is not extended in a global manner, it also guarantees that syntactic 
  conflicts cannot arise at link-time.

%There is also a flexible mechanism for explicit delimiters, for small DSLs, that we will discuss briefly later in the paper.

 % \emph{whitespace-delimited} at the top level, according to a particular strategy that we will describe.  Various forms of parentheses 
  %can also serve as delimiters.  Thus, indentation levels or
  %parenthesized expressions clearly delimit blocks that are governed by a particular 
  %language.  This makes the boundaries of each DSL clear to
  %the programmer and the compiler, enhancing usability and guaranteeing that 
  %ubtle conflicts cannot arise.

%\end{itemize}

\item Within this basic syntactic framework, we then propose a novel
type-directed dispatch mechanism: the \emph{expected type} of an expression, rather than an explicit keyword, 
%, rather than a keyword, 
determines which DSL grammar should parse the delimited block that generates that expression. That is, \emph{grammars are associated with types}. We will show that the  more common keyword-directed strategy arises as a special case of this strategy. 
\end{itemize}

This mechanism allows us to satisfy many of the criteria above, including safe composability, while still being quite expressive, as we will show with examples in the next section. We will continue by describing our approach in more detail (\S\ref{s:approach}), discuss ongoing research directions (\S\ref{s:discussion}), and conclude with related work (\S\ref{s:related}).
